Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-16 submission/manuscript.Rmd
intro edits

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

>>>>

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated germline-enriched gene list. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
